---
title: "Genome Annotation and the Interpretation of Gene Lists"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Differential Expression Analysis

```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"
```

```{r}
#Import metadata and take a peak.
colData <- read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
#Import countdata.
countData = read.csv(countFile, row.names = 1)
head(countData)
```

Hmm... remember that we need the `countData` and `colData` files to match up so we will need to remove that odd first column in `countData` namely `contData$length`.

```{r}
#Note we need to remove the odd first $length col.
countData <- as.matrix(countData[, -1])
head(countData)
```

Double-check that the colums in countData match the id calues in the column metadata file.
```{r}
colnames(countData)
```

```{r}
rownames(colData)
```

```{r}
all(colnames(countData) == rownames(colData))
```

This looks better but there are lots of zero entries in there so let's get rid of them as we have no data for these.

```{r}
countData = countData[rowSums(countData) != 0, ]
head(countData)
```

##Running DESeq2

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```

```{r}
dds
```

Next, get results for the HoxA1 knockdown versus control siRNA (remember that these were labeled as "hoxa1_kd" and "control_sirna" in our original `colData` `metaFile` input to DESeq, you can check this above and by running `resultsNames(dds)` command.

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```

```{r}
summary(res)
```

##Volcono plot

```{r}
plot( res$log2FoldChange, -log(res$padj) )

# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (-log(res$padj)) < 0.01 & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

##Adding gene annotation

Since we mapped and counted against the Ensembl annotation, our results only have information about Ensembl gene IDs. However, our pathway analysis downstream will use KEGG pathways, and genes in KEGG pathways are annotated with Entrez gene IDs. So lets add them as we did the last day.

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(countData), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(countData),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

#res$name = mapIds(org.Hs.eg.db,
                    #keys=row.names(res),
                    #keytype="ENSEMBL",
                    #column="NAMES",
                    #multiVals="first")

head(res, 10)
```

#Pathway Analysis
Here we are going to use the gage package for pathway analysis. Once we have a list of enriched pathways, we're going to use the pathview package to draw pathway diagrams, shading the molecules in the pathway by their degree of up/down-regulation.
```{r}
library(pathview)
```

##KEGG Pathways

The gageData package has pre-compiled databases mapping genes to KEGG pathways and GO terms for common organisms. `kegg.sets.hs` is a named list of 229 elements. Each element is a character vector of member gene Entrez IDs for a single KEGG pathway. (See also `go.sets.hs`). The `sigmet.idx.hs` is an index of numbers of signaling and metabolic pathways in `kegg.set.gs`. In other words, KEGG pathway include other types of pathway definitions, like "Global Map" and "Human Diseases", which may be undesirable in a particular pathway analysis. Therefore, `kegg.sets.hs[sigmet.idx.hs]` gives you the "cleaner" gene sets of signaling and metabolic pathways only.

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

The main `gage()` function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
head(keggres$less)
```

Now, let's try out the `pathview()` function from the pathview package to make a pathway plot with our RNA-Seq expression results shown in color. To begin with lets manually supply a `pathway.id` (namely the first part of the "hsa04110 Cell cycle") that we could see from the print out above.

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

Note how many of the genes in this pathway are perturbed (i.e. colored) in our results.

Now, let's process our results a bit more to automagicaly pull out the top 5 upregulated pathways, then further process that just to get the pathway IDs needed by the `pathview()` function. We'll use these KEGG pathway IDs for pathview plotting below.

```{r}
#Focus on top 5 upregulated pathways here for demo purposes only.
keggrespathways <- rownames(keggres$greater)[1:5]

#Extract the 8 character long IDs part of each string.
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

Finally, lets pass these IDs in keggresids to the `pathview()` function to draw plots for all the top 5 pathways.

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

Can you do the same procedure as above to plot the pathview figures for the top 5 down-regulated pathways?

```{r}
keggrespathwaysdown <- rownames(keggres$less)[1:5]

#Extract the 8 character long IDs part of each string.
keggresidsdown = substr(keggrespathways, start=1, stop=8)
keggresidsdown
```

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresidsdown, species="hsa")
```

#Gene Ontology (GO)

We can also do a similar procedure with gene ontology. Similar to above, `go.sets.hs` has all GO terms. `go.subs.hs` is a named list containing indexes for the BP, CC, and MF ontologies. Let’s focus on BP (a.k.a Biological Process) here.