---
title: "Introduction to Machine Learning for Bioinformatics"
author: "Julie Nguyen"
date: "October 24, 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#K-means Clustering
```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))

plot(x)
```

```{r}
k <- kmeans(x, center=2, nstart=20)
```

```{r}
k$cluster
```

```{r}
k$center
```

```{r}
plot(x, col=k$cluster)
points(k$centers, col="blue", pch=16)
```

#Hierarchical Clustering
The `hclust()` function requires a distance matrix as input. You can get this from the `dist()` function.

```{r}
# First we need to calculate point (dis)similarity
# as the Euclidean distance between observations
dist_matrix <- dist(x)

# The hclust() function returns a hierarchical
# clustering model
hc <- hclust(d = dist_matrix)

# the print method is not so useful here
hc 
```

```{r}
#Draw a dendogram.
plot(hc)
#The large jump in the height of the "crossbar" tells us that this tree is a good one.
abline(h=6, col="red")
cutree(hc, k=2) #Cuts tree into k groups.
cutree(hc, h=6) #Cuts tree based on height.
```

```{r}
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
```

```{r}
plot(x)
col <- as.factor(rep(c("c1","c2","c3"), each=50))
plot(x, col=col)
```

```{r}
#Create hierarchical clustering of the data.
hc <- hclust(dist(x))
#Draw a tree.
plot(hc)
#Cut the tree into 3 groups.
abline(h=2, col="red")
grps <- cutree(hc, k=3)
#Plot according to the number of groups.
plot(x, col=grps)
```

How many points in each cluster?
```{r}
table(grps)
```

Cross-tabulate (i.e. compare our clustering results with the known answer.)
```{r}
table(grps, col)
```

#Principal Component Analysis
```{r}
mydata <- read.csv("https://tinyurl.com/expression-CSV",
  row.names=1)
head(mydata)
```

How many genes are in this dataset?
```{r}
nrow(mydata)
```

Let's do PCA with the `prcomp()` function.
```{r}
pca <- prcomp(t(mydata), scale=TRUE)
attributes(pca)
```


```{r}
#Basic PC1 vs. PC2 plot.
plot(pca$x[,1], pca$x[, 2])

#Percent variance is often more informative to look at .
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

pca.var.per

#Scree plot.
barplot(pca.var.per, main="Scree Plot",
  xlab="Principal Component", ylab="Percent Variation")
```

Generate a PCA plot by grouping with colors.
```{r}
colvec <- colnames(mydata)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)"))

#Can also generate the same plot this way:
plot(pca$x[,1], pca$x[, 2],
      col=c("red", "red", "red", "red", "red",
            "blue", "blue", "blue", "blue", "blue"))
```

#PCA of UK food data
```{r}
x <- read.csv("UK_foods.csv")
dim(x)
```

Set the row names.
```{r}
#Assign the first column as the row names for the data set.
rownames(x) <- x[, 1]

#Redefine x by adding the row names to anything but the first column.
x <- x[, -1]

#The method above is dangerous because re-running this code chunk will change the dimensions of the dataset every time.
#Can also set row names with read.csv(file, row.names=1).

#Preview the first 6 rows.
head(x)
dim(x)
```

Use the more robust approach to setting row names.
```{r}
x <- read.csv("UK_foods.csv", row.names=1)
```

###Spotting major differences and trends.
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))

#Generate a barplot with stacked bars.
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

Generate all pairwise plots.
```{r}
pairs(x, col=rainbow(10), pch=16)

#What does it mean if a given point lies on the diagonal for a given plot?
#
#If a point lies on a diagonal, then the two countries have similar eating habits.
```

###PCA to the rescue!
```{r}
pca <- prcomp(t(x))
#t argument transduces (i.e. flips) the rows and columns. We need to do this because prcomp() expects observations to be rows and variables to be columns.

summary(pca)
#Proportion of Variance tells us how much each PC accounts for in the total 17-dimension variance. Cumulative Proportion will tell us how much each group of PC accounts for in the total 17-dimension variance.
```

Generate a plot of PC1 vs. PC2.
```{r}
plot(pca$x[, 1], pca$x[, 2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
```

