---
title: "Class07"
author: "Julie Nguyen"
date: "October 22, 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R functions revisited.

Source my functions from the last class.

```{r}
source("http://tinyurl.com/rescale-R")
```

Testing the **rescale()** and **rescale2()** functions.
```{r}
rescale(1:10)
```

```{r}
rescale(c(1, 10, 5, NA, 6))
```

rescale2() throws an error that is helpful and informative.
```{r}
rescale2(c(1:10, "Julie"))
```

rescale() throws an error that is not so helpful...
```{r}
rescale(c(1:10, "Julie"))
```

## A new function called both_na().
Write a function to find where there are NA elements in two input vectors.

First make some simple input where I know the answer.
```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3, 4)
```

Looked online and found the **is.na()** function
```{r}
is.na(x)
```

and the **which()** function tells me where the TRUE values are.
```{r}
which(is.na(x))
```

The AND function requries two input TRUE to give a TRUE output.
```{r}
is.na(x) & is.na(y)
```

Taking the **sum()** of TRUE FALSE vectors will tell me how many TRUE elements I have. This is my working snippet!!
```{r}
sum(is.na(x) & is.na(y))
```

Now turn it into a function.
```{r}
both_na <- function(x, y){
  sum(is.na(x) & is.na(y))
}
```

```{r}
both_na(x, y)
```

```{r}
x <- c(NA, NA, NA)
y1 <- c(1, NA, NA)
y2 <- c(1, NA, NA, NA)
```

```{r}
both_na(x, y2)
#Will yield the message "longer object length is not a multiple of shorter object length [1] 3"
#The number 3 in the error message is a result of R recycling the elements of x in order to fit the longer vector, y2.
```

Hypothesized that the number 3 in the error message is the return value of the length of the shorter vector. (But this is false.)
```{r}
x2 <- c(NA, NA)
both_na(x2, y2)
```

```{r}
x3 <- c(1, NA, NA, 1, NA, NA, 1)
y3 <- c(1, NA, NA, NA, NA, NA, NA)
```

```{r}
length(x)
length(y3)
```

Add a check for when inputs x and y are not the same length.
```{r}
both_na2 <- function(x, y){
  if(length(x) != length(y)){
    stop("Input x and y should be the same length!")
  }
    sum(is.na(x) & is.na(y))
}
```

Test if the updated both_na2() function works.
```{r}
both_na2(x, y3)
```

Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single alignment score.
```{r}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
```

```{r}
grade <- function(x) {
  if(any(is.na(x))) {
    warning("Student is missing an assignment!")
  }
  mean(x[- which.min(x)], na.rm = TRUE)
}
```

```{r}
hw <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
```

```{r}
apply(hw, 1, grade)
```
